Sampling Strategies in Decisions from Experience

### 2022-02-25 manuscript.Rmd
Update Simulation Study

```{r packages}
# load required packages
pacman::p_load(papaja,
               readr,
               tidyr, 
               dplyr,
               ggplot2,
               ggbeeswarm,
               scico)
```

@@

Now, more formally, let the sample of a prospect be a set of realizations of the respective random variable according to its distribution $P_X$.^[<!--Open footnote-->Note that this paper considers random samples, although it has been emphasized that in the natural world samples are "virtually never random" [@fiedlerBewareSamplesCognitiveecological2000, p. 660]. To account for non-random sampling, e.g., in the form of sampling biases [see @fiedlerBewareSamplesCognitiveecological2000] or dynamic information structures [see, e.g., @cohenEffectPerceivedPatterns2021; @plonskyRelianceSmallSamples2015], the i.i.d-assumption would need to be dropped.<!--Close footnote-->]
So, for the choice between $1, ..., k$ prospects, the inputs to DfE are generated by a set $S$ of stochastic processes
$$
S = \{X_{1t}, ... , X_{kt}\}
\tag{8}
$$
with $t \in \{1, ..., N\}$ denoting a sequence of independent and identically distributed (i.i.d.) random variables and $N$ denoting the length of the respective sequence.
The realizations that result from each stochastic process $s \in S$ may then be integrated into a sample of $N$ experienced outcomes or into multiple smaller samples.
The samples may then be used to make an uncertain inference about the latent prospects in (7), the sets of outcome-probability pairs respectively.

Of course, the laws of large numbers [LLN, see, e.g., @kolmogorovFoundationsTheoryProbability1950] apply to DfE and it follows that the accuracy of an inference from a sample about the properties of a latent prospect depends, ceteris paribus, on the size of the sample.
In fact, one explanation for the as-if underweighting of rare outcomes in DfE and the description-experience gap outlined in the introduction is that people rely on too small samples, which, due to the skewness of the binomial distribution, cause the relative frequencies with which rare outcomes occur in a sample to be smaller than their objective probabilities [@hertwigDecisionsExperienceEffect2004; @rakowBiasedSamplesNot2008; see also @hertwigDecisionsExperienceWhy2010].
The invoked argument is then that what may appear as if people were underweighting rare outcomes relative to what would be warranted on the basis of their objective probabilities---following the principle of mathematical expectation---may in fact be consistent with an as-if overweighting relative to what would be warranted on the basis of the experienced relative frequencies [@foxDecisionsExperienceSampling2006].
Yet, it seems that the small size of samples in DfE provides just one and only a partial explanation for the description-experience gap [@hauDescriptionexperienceGapRisky2008; @hauDecisionsExperienceStatistical2010; @ungemachAreProbabilitiesOverweighted2009; see also @wulffMetaanalyticReviewTwo2018].
More so, despite the LLN, the gap need not narrow at all, even for ever increasing sample sizes that are fully taken into account.
Particularly, this paper discusses and demonstrates in a simulation how different strategies of sequentially generating outcomes and integrating them into samples interact with the LLN to produce distinct choice patterns in DfE---including the as-if underweighting of rare outcomes---that tend to stabilize as the sample size increases.
Before so doing, however, the following section reviews how CPT can be used as a data model that captures systematic deviations from the principle of mathematical expectation in the parameter estimates of a *value* and a *weighting function*.

@@

```{r w-shapes, fig.cap="(ref:weighting)"}
wf <- tibble(gamma = seq(.1, 2, .1)) %>%
  expand_grid(delta = c(.1, .5, 1, 2, 5, 10),
              prob = seq(0, 1, .01)) %>%
  mutate(w = (delta*(prob^gamma))/((delta*prob^gamma)+(1-prob)^gamma))

ggplot(wf) +
  geom_line(aes(prob, w, group = gamma, color = gamma)) +
  geom_segment(x = 0, xend = 1, y =  0, yend = 1, linetype = "dashed", size = .2, color = "gray") +
  scale_color_scico(palette = "cork") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x = "p",
       y = "w(p)", 
       color = expression(gamma)) + 
  facet_wrap(. ~ delta, labeller = label_bquote(cols = delta: .(delta))) + 
  theme_minimal()
```

To begin with, the gray-dashed identity lines in Figure\ \@ref(fig:w-shapes) imply a linear weighting pattern.
That is, since the images of $w$, $w(p)$, are equal to the respective cumulative probabilities $p$, all cumulative decision weights $\pi$ that are derived according to (12) would be equal to the differences between the respective objective cumulative probabilities and therefore also to the objective probabilities of the outcomes for which the weights are determined.
Such a linear weighting pattern is implied by $\gamma = 1$ and $\delta = 1$. 
Now, both deviations from $\gamma = 1$ and $\delta = 1$ produce a nonlinear graphical shape.
That is, for $\gamma > 1$ (green graphs), the graphs take a S-shape---running below (above) the identity line for small (large) cumulative probabilities---which is more accentuated for larger deviations from $\gamma = 1$.
Importantly, the same *small* numerical difference between two cumulative probabilities $p$ then translates to varying differences between the respective images $w(p)$ depending on the interval on the cumulative probability scale. 
Specifically, because of the S-shaped curvature, the differences between the images $w(p)$ may be smaller than those between the respective cumulative probabilities in the lower and upper part of the cumulative probability scale, but may be considerably greater in the middle part.
As a consequence, the cumulative decisions weights are smaller than the respective objective probabilities for small-probability outcomes of either a low rank or a high rank, but they are larger for small-probability outcomes of a middle rank.^[<!--Open footnote-->Note that the distinction between outcomes of either low/high rank or middle rank is unnecessary for two-outcome prospects [see @tverskyAdvancesProspectTheory1992]. However, for finite many outcomes, the distinction is critical. More so, the fact that CPT may weigh outcomes of the same small probability differently, adds another reasons why the description-experience gap should not be equated with a reversed weighting pattern in CPT [see @hertwigConstructBehaviorGap2018]. I.e., neither could the gap then be interpreted in terms of an as-if underweighting (overweighting) of rare outcomes in DfE (DfD), nor would any of the potential contributors to the gap considered so far [see, e.g., @hertwigDescriptionexperienceGapRisky2009; @wulffMetaanalyticReviewTwo2018] treat some small-probability outcomes differently than others.<!--Close footnote-->]
Note that for all $\gamma < 1$ (blue graphs) the graph of $w$ takes an inverse S-shape and the entire weighting pattern is reversed.
Moreover, the shape of the weighting function, and thus the weighting pattern, depends strongly on $\delta$, which affects the overall elevation of the graph.
That is, for $\delta < 1$, the interval on the cumulative probability scale over which the weighting function runs below the identity line is greater than the interval over which the weighting function runs above the identity line; and this pattern reverses for $\delta > 1$.
More specifically, $\delta$ shifts across the entire cumulative probability scale the intervals within which small differences between objective cumulative probabilities translate to either smaller or greater cumulative decision weights.
In other words then, for any decrease (increase) in $\delta$, the underweighting of small probability outcomes extends to more outcomes at the lower (higher) end of the outcome range.

@@

@@

trial_length %>%
  ggplot(aes(x = s, y = med)) +
  geom_jitter(color = "#CECECE", size = 3) +
  geom_point(data = trial_length_s, aes(color = s), size = 3) +
  geom_path(data = trial_length_s, aes(color = s)) +
  scale_color_scico(palette = "buda") + 
  scale_x_continuous(breaks = seq(0, 1, .1)) + 
  scale_y_continuous(breaks = seq(0, 120, 10)) +
  labs(title = "Trial Length", 
       x ="Switching Probability",
       y = "Median Trial Length", 
       color="Switching Probability") + 
  theme_apa()
```

@@

```{r eval=FALSE}
cpt_long %>% 
  filter(parameter == "gamma") %>% 
  ggplot(aes(x=s, y = mean, color = s)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point(size = 2) +
  scale_color_scico(palette = "acton") + 
  scale_x_continuous(breaks = seq(0,1,.2)) +
  facet_grid(boundary~a) + 
  theme_minimal() 
```

>>>>

```{r packages}
# load required packages
pacman::p_load(papaja,
               readr,
               tidyr, 
               dplyr,
               ggplot2,
               viridis,
               scico,
               latex2exp)
```

@@

Now, more formally, let the sample of a prospect be a set of realizations of the respective random variable according to its distribution $P_X$.
That is, for the choice between $1, ..., k$ prospects, the inputs to DfE are generated by a set $S$ of stochastic processes
$$
S = \{X_{1t}, ... , X_{kt}\}
\tag{8}
$$
with $t \in \{1, ..., N\}$ denoting a sequence of independent and identically distributed (i.i.d.) random variables and $N$ denoting the length of the respective sequence [but see, e.g., @plonskyRelianceSmallSamples2015, where the i.i.d.-assumption is dropped].
So, given a stochastic process $s \in S$, the obtained realizations---i.e., the experienced outcomes---may be sequentially integrated into a sample of size $N$ or into multiple smaller samples.
Then, the samples so obtained can serve probabilistic inferences about the properties of the latent prospects in (7), the sets of outcome-probability pairs respectively.

Of course, the laws of large numbers [LLN, see, e.g., @kolmogorovFoundationsTheoryProbability1950] apply to DfE and it follows that the accuracy of inferences from samples to latent prospects depend, ceteris paribus, on the size of the samples.
In fact, one explanation for the as-if underweighting of rare outcomes in DfE and the description-experience gap outlined in the introduction is that people rely on too small samples, which, due to the skewness of the binomial distribution, cause the relative frequencies with which rare outcomes occur in the samples to be smaller than their objective probabilities [@hertwigDecisionsExperienceEffect2004; @rakowBiasedSamplesNot2008; see also @hertwigDecisionsExperienceWhy2010].
The invoked argument is then that what may appear as if people were underweighting rare outcomes relative to what would be warranted on the basis of their objective probabilities---following the principle of mathematical expectation---may in fact be consistent with an as-if overweighting relative to what would be warranted on the basis of the experienced relative frequencies [@foxDecisionsExperienceSampling2006].
Yet, it seems that the small size of samples in DfE provides just one and only a partial explanation for the description-experience gap [@hauDescriptionexperienceGapRisky2008; @hauDecisionsExperienceStatistical2010; @ungemachAreProbabilitiesOverweighted2009; see also @wulffMetaanalyticReviewTwo2018].
More so, despite the LLN, the gap need not narrow at all, even for ever increasing sample sizes that are fully taken into account.
Particularly, this paper discusses and demonstrates in a simulation how different strategies of sequentially generating outcomes and integrating them into samples interact with the LLN to produce distinct choice patterns in DfE---including the as-if underweighting of rare outcomes---that tend to stabilize as the sample size increases.
Before so doing, however, the following section reviews how CPT can be used as a data model that captures systematic deviations from the principle of mathematical expectation in the parameter estimates of a *value* and a *weighting function*.

@@

```{r w-shapes, fig.cap="(ref:weighting)"}

# prepare vector with cumulative probabilities
p <- tibble(p = seq(0, 1, .01))

# compute weighting function for different combinations of gamma and delta 
wf <- p %>% expand_grid(gamma = seq(.1, 2, .1), 
                        delta = c(.1, .5, 1, 2, 5, 10)) %>%
  mutate(wp = (delta*(p^gamma))/((delta*p^gamma)+(1-p)^gamma))

# labeller function for LaTeX math expressions in facet labels 
labeller_delta <- function(string) {
  TeX(paste("$\\delta=$", string, sep = ""))  
}

# plot
ggplot(wf, aes(p, wp, group = gamma)) +
  facet_wrap(~delta, labeller = as_labeller(labeller_delta, default = label_parsed)) + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x = expression(p),
       y = expression(w(p)), 
       color = expression(gamma)) + 
  theme_apa() + 
  geom_line(aes(color = gamma)) +
  scale_color_scico(palette = "berlin") + 
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed", size = 1) # identity line
```

To begin with, the gray-dashed identity lines in Figure\ \@ref(fig:w-shapes) imply a linear weighting pattern.
That is, since the images of $w$, $w(p)$, are equal to the respective cumulative probabilities $p$, all cumulative decision weights $\pi$ that are derived according to (12) would be equal to the differences between the respective objective cumulative probabilities and therefore also to the objective probabilities of the outcomes for which the weights are determined.
Such a linear weighting pattern is implied by $\gamma = 1$ and $\delta = 1$. 
Now, both deviations from $\gamma = 1$ and $\delta = 1$ produce a nonlinear graphical shape.
That is, for $\gamma > 1$, the graphs take a S-shape---running below (above) the identity line for small (large) cumulative probabilities---which is more accentuated for larger deviations from $\gamma = 1$.
Importantly, the same *small* numerical difference between two cumulative probabilities $p$ then translates to varying differences between the respective images $w(p)$ depending on the interval on the cumulative probability scale. 
Specifically, because of the S-shaped curvature, the differences between the images $w(p)$ may be smaller than those between the respective cumulative probabilities in the lower and upper part of the cumulative probability scale, but may be considerably greater in the middle part.
As a consequence, the cumulative decisions weights are smaller than the respective objective probabilities for small-probability outcomes of either a low rank or a high rank, but they are larger for small-probability outcomes of a middle rank.^[<!--Open footnote-->Note that the distinction between outcomes of either low/high rank or middle rank is unnecessary for two-outcome prospects [see @tverskyAdvancesProspectTheory1992]. However, for finite many outcomes, the distinction is critical. More so, the fact that CPT may weigh outcomes of the same small probability differently, adds another reasons why the description-experience gap should not be equated with a reversed weighting pattern in CPT [see @hertwigConstructBehaviorGap2018]. I.e., neither could the gap then be interpreted in terms of an as-if underweighting (overweighting) of rare outcomes in DfE (DfD), nor would any of the potential contributors to the gap considered so far [see, e.g., @hertwigDescriptionexperienceGapRisky2009; @wulffMetaanalyticReviewTwo2018] treat some small-probability outcomes differently than other small-probability outcomes.<!--Close footnote-->]
Note that for all $\gamma < 1$ the graph of $w$ takes an inverse S-shape and the entire weighting pattern is reversed.
Moreover, the shape of the weighting function, and thus the weighting pattern, depends strongly on $\delta$, which affects the overall elevation of the graph.
That is, for $\delta < 1$, the interval on the cumulative probability scale over which the weighting function runs below the identity line is greater than the interval over which the weighting function runs above the identity line; and this pattern reverses for $\delta > 1$.
More specifically, $\delta$ shifts across the entire cumulative probability scale the intervals within which small differences between objective cumulative probabilities translate to either smaller or greater cumulative decision weights.
In other words then, for any decrease (increase) in $\delta$, the underweighting of small probability outcomes extends to more outcomes at the lower (higher) end of the outcome range.

@@

```{r}
cols <- list(.default = col_double(),
             boundary = col_factor(),
             gamble = col_factor(),
             rare = col_factor(),
             agent = col_factor(),
             choice = col_factor())
choices_p <- read_csv("data/choices/choices.csv", col_types = cols) # piecewise integration

# Choices

# determine normative choice objective and experienced
fr_rates_p <- choices_p %>%
  mutate(norm = case_when(ev_ratio > 1 ~ "A", ev_ratio < 1 ~ "B"), # normative choice objective EV
         ev_ratio_exp = round(a_ev_exp/b_ev_exp, 2),
         norm_exp = case_when(ev_ratio_exp > 1 ~ "A", ev_ratio_exp < 1 ~ "B")) # normative choice experienced EV / mean

# objective
fr_rates_p_o <- fr_rates_p %>%
  filter(!is.na(norm)) %>%
  group_by(s, boundary, a, rare, norm, choice) %>%
  summarise(n = n()) %>%
  mutate(rate = round(n/sum(n), 2), # false response rates
         type = case_when(norm == "A" & choice == "B" ~ "Safe", norm == "B" & choice == "A" ~ "Risky")) %>%
  ungroup() %>%
  filter(!is.na(type)) %>% # remove correct responses
  select(-c(norm, n)) %>% 
  mutate(norm = "Latent")

# experienced
fr_rates_p_e <- fr_rates_p %>%
  filter(!is.na(norm_exp)) %>%
  group_by(s, boundary, a, rare, norm_exp, choice) %>%
  summarise(n = n()) %>%
  mutate(rate = round(n/sum(n), 2), # false response rates
         type = case_when(norm_exp == "A" & choice == "B" ~ "Safe", norm_exp == "B" & choice == "A" ~ "Risky")) %>%
  ungroup() %>%
  filter(!is.na(type)) %>%  # remove correct responses
  select(-c(norm_exp, n)) %>% 
  mutate(norm = "Sample")

# merge data frame
fr_rates_p <- bind_rows(fr_rates_p_o, fr_rates_p_e) %>% 
  mutate(rare = case_when(rare == "none" ~ "Middle",
                          rare == "attractive" ~ "High", 
                          rare == "unattractive" ~ "Low"))

labeller_a <- function(string) {
  TeX(paste("$\\a=$", string, sep = ""))  
}

fr_rates_p %>% 
  ggplot(aes(s, rate)) +
  geom_point(aes(color = norm, shape = type)) +
  facet_grid(rare~a, labeller = labeller(rare = label_value, a = as_labeller(labeller_a, default = label_parsed))) + 
  scale_color_scico_d(palette = "berlin")+ 
  scale_x_continuous(limits = c(0, 1), breaks = c(0, .5, 1)) + 
  scale_y_continuous(limits = c(0, 1), breaks = c(0, .5, 1)) +
  labs(x = "s",
       y = "False Response Rate", 
       color = "Normative\nEV",
       shape = "False\nResponse") +
  theme_apa()
```

@@

trial_length %>%
  ggplot(aes(x = s, y = med)) +
  geom_jitter(color = "#CECECE", size = 3) +
  geom_point(data = trial_length_s, aes(color = s), size = 3) +
  geom_path(data = trial_length_s, aes(color = s)) +
  scale_color_scico(palette = "acton") + 
  scale_x_continuous(breaks = seq(0, 1, .1)) + 
  scale_y_continuous(breaks = seq(0, 120, 10)) +
  labs(title = "Trial Length", 
       x ="Switching Probability",
       y = "Median Trial Length", 
       color="Switching Probability") + 
  theme_apa()
```

@@

```{r eval=FALSE}
cpt_long %>% 
  filter(parameter == "delta", a!= 1) %>% 
  ggplot(aes(x=s, y = mean, color = s)) +
  geom_errorbar(aes(ymin=`2.5%`, ymax=`97.5%`)) + 
  geom_point(size = 2) +
  scale_color_scico(palette = "acton") + 
  scale_x_continuous(breaks = seq(0,1,.2)) +
  facet_grid(boundary~a) + 
  theme_minimal() 
```
